Gravity disturbance or gravity anomaly?

### 2018-04-16 manuscript/letter.tex
include new figures and define the unit vectors

%Extra packages
\usepackage{graphicx}

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a geocentric and rigid ellipsoid 
of revolution, which has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth (Figure \ref{fig:fig1}).
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
This model is called as \textit{normal Earth} 
\citep{heiskanen-moritz1967,vanicek1987,hofmann-wellenhof-moritz2005,
torge2012}.

It is worth noting that, although the normal Earth has the
same total mass (including the atmosphere) of the Earth,
its internal density distribution is unknown.

The only condition imposed on its internal density
distribution is that it produces a gravity field
having a particular equipotential which coincides
with its limiting surface.

For convenience, we denote any density distribution 
satisfying this condition as a \textit{normal density distribution}.

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
simply \textit{normal gravity}.

The normal Earth gives rise to the \textit{geodetic coordinate system}.
In this coordinate system, the position of a point $P$
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude}
$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).

The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this \textit{geocentric Cartesian coordinate system}
is called meridian plane (gray plane in Figure \ref{fig:fig1}).

@@

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/geodetic_coordinates.png}
    \caption{Schematic representation of the geodetic coordinate
    system defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the mean Earth's rotation axis, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, the geodetic latitude 
    $\varphi$ and longitude $\lambda$. The Earth's origin is represented 
    by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
    $P^{\prime}$ represents the projection of $P$ on the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
    In (b), the point $Q$ represents the projection of $P$ on the 
    ellipsoid surface.}
  \label{fig:fig1}
\end{figure}

>>>>

%Extra packages
\usepackage{graphicx}
\usepackage{amsmath}

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a geocentric and rigid ellipsoid 
of revolution, which has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth \citep{heiskanen-moritz1967,
vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
Here, we follow \citep{torge2012} and call this model as
\textit{normal Earth}.

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
simply \textit{normal gravity}.
In geodesy, any model used to represent the normal gravity field
can be arbitrarily defined for the only purpose
of keeping the difference from the actual gravity field as small 
as possible \citep{vanicek1987}.

It is worth noting that, although the normal Earth has the
same total mass (including the atmosphere) of the Earth,
its internal density distribution is unknown.
The search for physically meaningful mass distributions 
that generate a required normal gravity field
has geophysical rather than geodetic motives \citep{marussi1974}.
The only condition imposed on its internal density
distribution is that it produces a gravity field
having a particular equipotential which coincides
with its limiting surface.
For convenience, we denote any density distribution 
satisfying this condition as a \textit{normal density distribution}.

The normal Earth gives rise to the \textit{geodetic coordinate system}
\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
In this coordinate system, the position of a point $P$
is defined by the \textit{geometric height} $h$, 
\textit{geodetic latitude}
$\varphi$ and \textit{longitude} $\lambda$ (Figure \ref{fig:fig1}).
Geodetic coordinates $(h, \varphi, \lambda)$ can be easily 
converted into geocentric Cartesian coordinates $(X, Y, Z)$
(Figure \ref{fig:fig1}).
The plane containing the point $P$, the axis $Z$ and
the origin $O$ of this \textit{geocentric Cartesian coordinate system}
is called meridian plane (gray plane in Figure \ref{fig:fig1}).

At a given point $(h, \varphi, \lambda)$, there are three 
unit vectors (Figure \ref{fig:fig1}) given by \citep{soler1976}:
\begin{equation}
\begin{split}
\hat{\mitbf{u}} &= 
\begin{bmatrix}
\cos\varphi \, \cos\lambda \\
\cos\varphi \, \sin\lambda \\
\sin\varphi
\end{bmatrix} \\
\hat{\mitbf{v}} &= 
\begin{bmatrix}
-\sin\varphi \, \cos\lambda \\
-\sin\varphi \, \sin\lambda \\
\cos\varphi
\end{bmatrix} \\
\hat{\mitbf{w}} &= 
\begin{bmatrix}
-\sin\lambda \\
\cos\lambda \\
0
\end{bmatrix}
\end{split} \: ,
\label{eq:unit-vectors}
\end{equation}
Notice that these unit vectors are mutually orthogonal.
The normal gravity vector $\mitbf{\gamma}_{P}$ at a point $P$
is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ at $P$.

\section{Gravity disturbance}

It is worth noting that, by definition, 
the centrifugal component of the normal gravity field is
equal to the centrifugal component of the Earth's gravity
field if they are evaluated at the same point.
Then, the differences between the gravity vector
(corrected from non-gravitational effects) 
and the normal gravity vector, at the same point, represents a purely 
gravitational and consequently harmonic disturbing field.
This gravitational disturbance 
is caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called ``anomalous masses" (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
``density anomalies" (e.g., \citeauthor{forsberg1984}, \citeyear{forsberg1984})
or ``gravity sources" (e.g., \citeauthor{blakely1996}, 
\citeyear{blakely1996}). Here, we opted for using the last term.

The difference between the observed gravity and the
normal gravity, at the same point, is called gravity disturbance
\citep{hofmann-wellenhof-moritz2005}.
Notice that the gravity disturbance is not equivalent to the
magnitude of the difference between the gravity vector
and the normal gravity vector at the same point.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

The gravity anomaly is the commonly used quantity in applied 
geophysics. It is defined as the difference
between the gravity on the geoid and the normal gravity on the ellipsoid,
both at the same geodetic latitude and longitude.
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function in space 
\citep{barthelmes2013}.
Different gravity anomalies can be calculated, depending on the
corrections applied to them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
These corrections are usually called gravity reductions.
For example, the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The last one is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

There was a certain lack of comprehension regarding the
geophysical meaning of gravity anomalies until the
mid 90's.
As properly pointed out by \citet{chapin1996} at that time, 
``although the corrections which bring about a Bouguer 
gravity anomaly are well established, the reasons for doing
them are not well understood. One cause of this common 
misunderstanding is that the subject has been poorly presented in
many of the basic texts".
In his seminal book, \citet{blakely1996} brought some light
on the geophysical meaning of gravity anomalies from the 
perspective of applied geophysics. \citet{blakely1996} correctly
defined gravity sources as density contrasts between the actual
internal density distribution of the Earth and the internal density
distribution of the normal Earth.
However, he did not stress that, by removing the normal gravity 
evaluated on the ellipsoid from the gravity measured 
on the Earth's surface, the remaining disturbing field will reflect 
not only the effect produced by the gravity sources, but also a
small combination of gravitational and centrifugal effects.
This additional, non-harmonic and undesired effect is 
simply due to the calculation of the normal gravity at a point
other than that were the gravity is measured.

\section{Mathematical description of the gravity disturbance in a local coordinate system}

In a local- or regional-gravity study, 
geophysicists commonly use a topocentric Cartesian coordinate system
with origin at a point $P$ and axes $x$, $y$ and $z$ defined
by the unit vectors $\hat{\mitbf{v}}_{P}$,
$\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
(equation \ref{eq:unit-vectors}), respectively
(Figures \ref{fig:fig1} and \ref{fig:fig2}).
In this coordinate system, the observed gravity vector
$\mitbf{g}_{i}$, at a point $(x_{i}, y_{i}, z_{i})$, 
$i = 1, ..., N$, can be represented by
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \Delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\Delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and a
disturbing gravitational attraction produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

By definition, the gravity disturbance $\delta g_{i}$,
at the point $(x_{i}, y_{i}, z_{i})$, is given by 
\citep{hofmann-wellenhof-moritz2005}:
\begin{equation}
\delta g_{i} =  g_{i} - \gamma_{i} \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $g_{i} = \| \mitbf{g}_{i} \|$ and 
$\gamma_{i} = \| \mitbf{\gamma}_{i} \|$ are,
respectively, the observed gravity and the normal gravity at the
point $(x_{i}, y_{i}, z_{i})$.
Fortunately, the condition 
$\gamma_{i} \gg \| \Delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
By combining this condition and the definition of observed gravity vector
(equation \ref{eq:gravity-vector}), we can approximate the observed gravity
$g_{i}$ by a first order Taylor's expansion as follows 
\citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \Delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$ and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(equation \ref{eq:unit-vectors}) evaluated at the point $(h_{i}, \varphi_{i}, \lambda_{i})$, which is obtained by transforming $(x_{i}, y_{i}, z_{i})$
from the topocentric coordinate system (Figure \ref{fig:fig2}) to the
geodetic coordinate system (Figure \ref{fig:fig1}).

PAREI AQUI

This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (equation \ref{eq:gobs-approx}) 
coincides with the $z$ axis
of the local Cartesian coordinate system defined
at the beginning of this section. Consequently,
by using the approximation defined in equation \ref{eq:gobs-approx}, 
the gravity disturbance (equation \ref{eq:gravity-disturbance}) can be 
rewritten as follows \begin{equation}
\delta g^{o}_{i} \approx \hat{\mitbf{z}}^{\top} \Delta \mitbf{g}^{o}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{z}}^{\top} = \left[ \, 0 \; 0 \; 1 \right]$.
According to equation \ref{eq:gravity-disturbance-approx},
the gravity disturbance $\delta g^{o}_{i}$ (equation \ref{eq:gravity-disturbance})
represents the vertical component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a homogeneous gravity source can be represented 
by the following harmonic function
\begin{equation}
d^{o}_{i} = k_{g} \, G \, \rho \, \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $G$ is the Newtonian constant of gravitation
(in $m^{3} / (kg \, s^{2})$),
$k_{g} = 10^{5}$ is a constant factor
transforming from $m/s^{2}$ to milligal (mGal),
and $\partial_{z} \phi_{i}$ is a harmonic function representing the first 
derivative, evaluated at the observation point $(x_{i},y_{i},z_{i})$, 
$i = 1, ..., N$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int \frac{1}{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv
\label{eq:phi}
\end{equation}
with respect to the variable $z$. The 
integral is conducted over the coordinates $x^{\prime}$, $y^{\prime}$ 
and $z^{\prime}$ within the volume $v$ of the gravity source.
This equation can be easily generalized for the case
of multiple gravity sources.

\section{Gravity disturbance versus gravity anomaly}

Almost all interpretation techniques assume, implicitly or
directly, that the gravity data is harmonic (e.g., 
upward/downward continuation, 
data processing with equivalent layer,
conversions between gravity and magnetic data,
computation of vertical derivatives via Fourier and Hilbert
transforms).

Consequently, they implicitly or directly assume that
the gravity data approximates the gravitational disturbance.

Almost all forward modelling techniques compute
the vertical component of the gravitational attraction 
exerted by the geological bodies at the observation points.

Hence, almost all geophysicists implicitly compute 
the vertical component of the gravitational disturbance.

The gravity anomaly is defined as the difference between 
the gravity $\| \mitbf{g}_{P} \|$, at a point $P$ on the
Geoid, and the normal gravity $\| \mitbf{\gamma}_{Q} \|$,
on the reference elipsoid, where $P$ and $Q$ have the same
geodetic latitude and longitude.

Consequently, the gravity anomaly is a function of
the geodetic latitude and longitude only and cannot
be calculated at arbitrary heights.

Gravity anomalies require the computation of gravity
within the topographic masses, where the gravitational
disturbance is not harmonic.

@@

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/cartesian-geodetic-systems.png}
    \caption{Schematic representation of the geodetic coordinate
    system defined by an oblate ellipsoid with semi-minor axis $b$, 
    coincident with the mean Earth's rotation axis, and a semi-major
    axis $a$. In this coordinate system, the position of a point is
    determined by the geometric height $h$, the geodetic latitude 
    $\varphi$ and longitude $\lambda$. 
    The Earth's center of mass is represented 
    by $O$, $P$ represents a point $(h, \varphi, \lambda)$ and 
    $P^{\prime}$ represents the projection of $P$ on the plane $XY$ 
    (Equatorial plane). The plane containing $O$, $P$ and 
    $P^{\prime}$ is represented by the gray triangle in (a) and (b). 
    The unit vectors 
    $\hat{\mitbf{u}}_{P}$, $\hat{\mitbf{v}}_{P}$ and 
    $\hat{\mitbf{w}}_{P}$ define three mutually orthogonal 
    directions at each point $P$.
    In (b), the point $Q$ represents the projection of $P$ on the 
    ellipsoid surface, at the same latitude $\varphi$ and 
    longitude $\lambda$.}
  \label{fig:fig1}
\end{figure}

\begin{figure}
    \includegraphics{figures/local-system.png}
    \caption{(a) Schematic representation of the gravity vector
    $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
    disturbing gravitational attraction $\Delta \mitbf{g}_{P}$
    (equation \ref{eq:gravity-vector}) and the unit vector 
    $\hat{\mitbf{u}}_{P}$ (equation \ref{eq:unit-vectors}) 
    at the point $P$ and also the normal gravity vector
    $\mitbf{\gamma}_{Q}$ at the point $Q$ on the reference ellipsoid.
    At the point $P$ the normal gravity vector $\mitbf{\gamma}_{P}$ 
    is opposite to the unit vector $\hat{\mitbf{u}}_{P}$ and the 
    normal gravity $\mitbf{\gamma}_{Q}$ at the point $Q$ is opposite
    the another unit vector $\hat{\mitbf{u}}_{Q}$, which is
    close to $\hat{\mitbf{u}}_{P}$.
    (b) Schematic representation of a topocentric Cartesian 
    coordinate system with origin at a point $P$ and $x$, $y$ and $z$ 
    axes defined by the unit vectors $\hat{\mitbf{v}}_{P}$, 
    $\hat{\mitbf{w}}_{P}$ and $-\hat{\mitbf{u}}_{P}$ 
    (Figure \ref{fig:fig1}), respectively.}
  \label{fig:fig2}
\end{figure}
